Efficient and Scalable Entity Resolution with Fast Beta Linkage (fabl)

### 2021-06-20 methods_paper/fabl-paper.tex
fabl method paper first draft

\title{Efficient and Scalable Entity Resolution with Fast Beta Linkage  (fabl)}
\author{Brian Kundinger\textsuperscript{a} \and
  Jerome Reiter\textsuperscript{a} \and 
  Rebecca C.~Steorts\textsuperscript{b}}
\date{
  \textsuperscript{a}Department of Statistical Science, Duke University\\
  \textsuperscript{b}Department of Statistical Science and Computer Science, Duke University\\Principal Mathematical Statistician, United States Census Bureau\\
  \today}

@@

In this paper, we propose an extention to the \emph{Beta Record Linkage}
(\texttt{BRL}) method proposed proposed by \citep{sadinle2017} for this very end.  We increase the speed of the \texttt{BRL} procedure through the modification to the model specification that allows for parallel computing of the linkage parameter, and use hashing techniques to hasten calculations and reduce computational complexity. Additionally, we introduce storage efficient indexing (SEI), and a data partitioning approach to make the method amenable to large linkage tasks. To recognize the lineage from the original \texttt{BRL} method, we name our method \emph{Fast Beta Linkage} (\texttt{fabl},
pronounced ``fable'')

@@

\paragraph{Fellegi and Sunter Method}
Most record linkage techniques are derived from the seminal 1969 paper
by Fellegi and Sunter: ``A Theory for Record Linkage''. The defining
characteristic of their approach was to transform the sets of records,
which often contain text data that is difficult to model, into sets of
comparison vectors governed by parameters that can be more easily
estimated. Concretely, if files \(A\) and \(B\) have \(n_A\) and \(n_B\)
records respectively, and if the files share \(F\) fields in common upon
which to base the linkage, the Fellegi and Sunter approach generates an
\(n_A n_B \times F\) matrix \(\Gamma\), which contains similarity scores
between each pair of records across datasets. We say \(\gamma_{ij}\) is
the comparison vector for record \(i \in A\) and record \(j \in B\),
with \(\gamma_{ij}^f \in \{1, \ldots, L_f\}\) providing their similarity
score on the \(f^{th}\) field. For ease of modeling and computation, we
restrict these similarity scores to be discrete, ordinal variables, and
the construction of these is left to the modeler. We adopt the
convention that \(\gamma_{ij}^f = 1\) corresponds to the highest level
of agreement and \(\gamma_{ij}^f = L_f\) corresponds to the lowest. It
is common to use use binary 1-2 variables to indicate exact matching,
and 1-2-3 variables to provide an option for partial matching. For text
data, we calculate similarity based on Levenstein distance or some other
text similarity score, and bin these scores to integers for use in the
model. Jaro (1989) provided guidelines for appropriate binning for the
comparison vectors, but details of implementation are generally context
specific.

The likelihood in the original Fellegi Sunter model is a mixture model
through which each record pair is independently classified as a match or
nonmatch. This independent classification often leads to sets of matches
that break expectations of transitivity, which is undesirable. Although
Jaro devised an procedure to reduce a set of conflicting matches to the
mostly likely set of one-to-one matchings, the abundance of false
matches can bias the estimation of other parameters of the model.

@@

\paragraph{Deterministic Blocking}

\paragraph{Other Approach (Will probably remove)}Outside of the Fellegi-Sunter framework, some researchers have developed
methods that model the data directly, not a derived set of comparison
vectors. As an early example, Steorts et al 2016 produced
\texttt{blink}, which had the advantage of a computational complexity
that grew linearly with the data (as opposed to quadratically within
Fellegi Sunter), but was plagued with slow mixing times and was unable
to incorporate text data. Later, Steorts et al 2020 improved on this
method with \texttt{dblink}, which was able to simulate text data within
its Gibbs sampler by drawing from an empirical distribution, and
remedied its slow mixing through a probabilistic blocking approach.
However, this method remains computationally intensive and requires
distributed computing to manage even moderate linkage problems in
acceptable time. Because these methods fall outside of the
Fellegi-Sunter framework, we do not consider them further in this paper.

@@

Here the reader should the relationship between our proposed model and
that of Sadinle 2017. In his model, Sadinle constructs a prior
distribution on the entire \(\mathbf{Z}\) vector, which induces a Gibbs
sampler that strictly enforces one-to-one matching. In particular, this
sampler removes previously matches records from the set of candidate
records when sampling \(Z_j\), creating a dependency that makes the
sampler \emph{inherently serial}. We however use independent priors for
each \(Z_j\), creating a sampler that is \emph{perfectly
	parallelizable}, allowing for significant computational gains. More
importantly, since only the agreement pattern of \(Z_j\) is used for
calculations within the Gibbs sampler, and not the particular record
label, we can conduct this sampling only at the level of the unique
agreement patterns, offering even more computational savings. In doing
so however, we thereby weaken the one-to-one requirement from
\texttt{BRL}; our sampler does ensure that each record in \(B\) can be
matched to at most one record in \(A\), but allows for the possiblity
that multiple records in \(B\) match to the same record in \(A\). We
explore the ramifications of this distinction with the El Salvador case
study.

@@

In order to make fair comparisons against the Sadinle 2017 model, we
integrate over the posterior of \(\lambda\) and rearrange terms to
produce the final full conditional:

@@

For a more indepth explanation of this function and the induced Bayes
estimate, see Sadinle 2017.

@@

\section{Efficient and Scalable Implementation}
\label{sec:efficiency}
Broadly speaking, we increase our computational efficiency by
recognizing that record pairs contribute to posterior calculations only
through the agreement pattern of the \(\gamma_{ij}\) vector. Let
\(\mathcal{H}\) be the set of unique agreement patterns in the data, let
\(P = |\mathcal{H}|\) denote the total number of unique agreement
patterns. Note that \(P\) is bounded above by \(\prod_{f=1}^F L_f\), and
that this bound does not scale with \(n_A\) or \(n_B\). Prior to
processing the data, we identify all \(P\) patterns in \(\mathcal{H}\)
and enumerate them \(h_1, \ldots, h_P\). When the \((i,j)\) pair
exhibits the \(p^{th}\) agreement pattern, we say \((i,j) \in h_p\).
Wherever possible, we conduct calculations over these \(P\) agreement
patterns rather than the \(n_A \times n_B\) record pairs.

@@

The classic Fellegi Sunter method represents the \(\gamma_{ij}\)
comparison vector as vector of length \(F\), with each component
\(\gamma_{ij}^f\) taking on values in \(\{1, \ldots, L_f \}\). To ease
computations, we instead represent the comparison as a concatenation of
\(F\) many binary indicator vectors of lengths \(L_f\). For example, if
\(L_1 = L_2 = 2\) and \(L_3 = 3\), then \(\gamma_{ij} = (2, 1, 3)\)
under the classical framework becomes
\(\gamma_{ij} = (0, 1, 1, 0, 0, 0, 1)\) under our framework. This is a
bijective transformation that does not change the meaning of the data,
but this representation will ease calculations and posterior updates.

@@

Laslty, for large data, we can partition the two datasets \(A\) and
\(B\) into smaller blocks \(\{A_m\}\) and \(\{B_m\}\) for more
manageable computations. On a single machine, we can read-in data
sequentially, conduct hashing, compress information through SEI, and
delete the original data from memory before continuing with the next
chunk of data. With multiple cores or multiple machines, this can be
done in parallel. Thus the combination of hashing, SEI, and partitioning
allows us to conduct linkage tasks over much larger datasets.

@@

Lastly, we recognize that all posterior updates are governed by the
agreement patterns of the record pairs rather than the record labels
themselves. Thus we complete the entire Gibbs procedure first at the
level of the \(P\) agreement patterns with the first equation above.
After, we can back-fill the records corresponding to the agreement
patterns through the second equation. Sampling uniformly is
computational simple even for large sets of candidate records, but this
step can also be parallelized when working with large data.

@@

\[\text{Recall} = \frac{\text{Matches Correctly Identified}}{\text{True Matches in Data}}\]
\[\text{Precision} = \frac{\text{Declared Matches}}{\text{True Matches in Data}}\]
\[\text{F-Measure} = 2\left(\frac{\text{Recall} \times \text{Precision}}{\text{Recall} + \text{Precision}}\right)\]

@@

		\includegraphics{../notes/figures/sadinle_sim_plot} 
	}

@@

To demonstrate speed, we generate comparison vectors from pre-specified
distributions so that we can easily increase the size of the linkage
problem. Distributions are meant to emulate the behavior of similarity
scores across first name, last name, and day, month. For simplicity, we
consider only exact matching, so a vector (1, 0) corresponds to
agreeement and (0, 1) to disagreement. We simulate these data for
different values of \(n_A\) and \(n_B\), and compare the run-time of
\texttt{fabl} against \texttt{BRL}. Note that the number of unique
patterns \(P\) is bounded above by \(2^5 = 32\), a bound which is
consistently attained in the larger simulations.

@@

\begin{table}[ht]
	\centering
	\begin{tabular}{rrr}
		\hline
		& m & u \\ 
		\hline
		fname & (0.95, 0.05) & (0.01, 0.99) \\ 
		lname & (0.95, 0.05) & (0.01, 0.99) \\ 
		day & (0.95, 0.05) & (0.03, 0.97) \\ 
		month & (0.95, 0.05) & (0.08, 0.92) \\ 
		year & (0.95, 0.05) & (0.07, 0.93) \\ 
		\hline
	\end{tabular}
\end{table}

@@

	{\centering \includegraphics{../notes/figures/sadinle_speed_plot2}

@@

	\caption{Runtime for BRL and fabl to run 1000 Gibbs iterations, including hashing step for fabl, for increasing values of both $n_A$ and $n_B$}\label{fig:speed1}
\end{figure}

@@

	{\centering \includegraphics{../notes/figures/speed_plot_fixed_nB}

@@

	\caption{Runtime for BRL and fabl to run 1000 Gibbs iterations, including hashing step for fabl, with increasing $n_A$, and $n_B$ fixed at 500}\label{fig:speed2}
\end{figure}

@@

We now demonstrate the power of our method through two case studies. The first is a study of documented identifiable deaths (DID) from the El Salvadoran Civil War. Though the data files used here are small, this study shows how the computational complexity of \texttt{fabl} depends on the number of unique agreement patterns, and how significant computational gains can be acheived by simplifying the construction of the comparison vectors. Secondly, the case study reveals the impact of independently sampling $Z_j$ rather than strictly enforcing one-to-one matching as done by \texttt{BRL}. The second case study is the National Long Term Care Survey (NLTCS), a much larger linkage tasks which \texttt{BRL} is unable to complete but which we complete with \texttt{fabl} with relative ease.

\paragraph{El Salvadoran Civil War}The country of El Salvador was immersed in civil war from 1980 to 1991,
and throughout the time, several organizations attempted to document
casualties of the conflict. When estimating the total number of
casualities, one cannot simply sum the numbers recorded by each
organization, as it is likely that the same individuals are recorded in
multiple casuality lists. To obtain a more accurate estimate of the
casualties then, we follow Sadinle's 2017 paper and link records of El
Salvadoran civilian casualties from two sources: El Rescate - Tutela
Regal (ERTL) and the Salvadoran Human Rights Commission (CDHES, by its
acronym in Spanish). The ERTL dataset consists of digitized reports that
had been published throughout the conflict. The CDHES dataset consists
of casualties that had been reported directly to the organization, and
later digitized.

@@

We only utilized records with nonmissing entries for given and last
name, results in \(n_A = 4420\) files in CHDES and \(n_B = 1323\) files
in ERTL. The names were standardized to account for common misspellings
in the Spanish language, and then compared using a modified Levenstein
distance to account for the fact that second names are often ommited.
Place of birth is recorded by municipality and department within that
municipality; however, since department was missing in 95\% of records
in CHDES and 80\% of records in ERTL, we excluded it from our linkage
process. Thus we conduct linkage using given name, last name,
municipality, and day, month, and year of death. We again use flat
priors for the \(\mathbf{m}\) and \(\mathbf{u}\) parameters.

@@

	{\centering \includegraphics{../notes/figures/el_salvador/m_posterior_smallP}

@@

To mirror the original implementation, we constructed the comparison
vectors using 4 levels of agreement for each field, according to the
thresholds provided in Figure XX. This took 422 seconds for our proposed
method and 240 for \texttt{BRL}. However, we observed that posterior
distributions of several levels of the \(\mathbf{m}\) and \(\mathbf{u}\)
parameters were nearly identical, and that of the
\(4^5 \times 2 = 2048\) possible agreement patterns, only 1173 are
realized in the data. This leads us to believe that such high number of
aggreement levels creates unnecesary distictions in the data and makes
the comparison vectors less interpretable. Therefore we re-ran our
analysis with fewer agreement levels for each field (see Figure XX), and
obtained analogous results. With 216 possible agreement patterns, 159
were realized in the data, and our proposed method became much faster,
finishing in 124 seconds. Meanwhile \texttt{BRL} took 239 seconds,
relatively unchanged from the first implementation. This demonstrates
the way that the computational complexity of our method depends on the
number of unique agreement patterns, and how significant computional
gains can be made by simplifying the construction of the comparison
vectors. We also note that estimates of the \(\mathbf{m}\) and
\(\mathbf{u}\) paramters under each method are very similar, as shown in
Figure~\ref{fig:m-and-u}.

\subsection{Violations of One-to-One Matching}
\label{violations-of-one-to-one-matching}

Figure~\ref{fig:overlap-plot} shows the posterior distribution for \(D\), the number of
duplicates found across file for both \texttt{fabl} and \texttt{BRL}. We
see that \texttt{fabl} consistently overmatches within each Gibbs
iteration when compared to \texttt{BRL}, which is to be expected because
\texttt{BRL} explicitly prevents matches that violate one-to-one
matching throughout the entire sampler. Most of such matchings are due
to the randomness in the sampling procedure, and they occur sporadically
throughout the sampler in such a way that does not measurably influence
the eventual Bayes estimate.

@@

	{\centering \includegraphics{../notes/figures/el_salvador/overlap_distribution_smallP_bayes}

@@

Overall, both methods presented similar results. \texttt{fabl} yielded
an initial Bayes estimate of 195 matches found across files, and after
resolving matches that violated one-to-one requirements, yeilded 180
matches across files. This is acceptably close to the 181 matches found
by \texttt{BRL}. The main reason for this discrepancy is the difference
in how each model handles situations in which one file in \(B\) has
multiple plausible matches in \(A\).

@@

	{\centering \includegraphics{../notes/figures/el_salvador/rosa_records}

@@

	{\centering \includegraphics{../notes/figures/el_salvador/bad_mixing}

@@

Figure XX shows the values that \(Z_{825}\) and \(Z_{826}\) take on
throughout the Gibbs sampler, and demonstrates how each method handles
this situation. Under \texttt{fabl}, both records in \(B\) match to the
same record in \(A\) throughout the Gibbs process, creating consistent
violations of one-to-one matching. Under \texttt{BRL}, the Gibbs process
creates one matching configuration stays there for a while. However, if
one pair ``unmatches,'' then the other record has a chance to latch on.
Then, the Gibbs process is stuck with that matching status for a while,
resulting in a Gibbs process with poor mixing. Additionally,
\texttt{fabl} allows the modeler to inspect records with multiple
plausible matches, and if they desire, to then choose the record pairing
with the highest posterior probability. \texttt{BRL} in contrast, in
strictly enforcing one-to-one matching throughout the sampler, can lead
to situations where none of the plausible matches reach the threshold to
be identified through the Bayes estimate.

\paragraph{National Long Term Care Survey} The National Long Term Care Survey is a longitudinal study to track changes in health among Medicare recipients. It was conducted approximately every five years, and we concern ourselves here with linking records from years 1982 and 1989. The data only contains day, month, and year of birth, sex, state the individual lives in, and regional office. State and regional office have been anomynized, presented as numbers instead of their actual values. We construct the comparison vectors using only binary exact matching for each field, since there is no straightfoward way to incorporate partial matching in this context. The 1982 file contains $n_A = 20485$ records, 1989 file contains $n_B = 17466 files$, and the six linkage fields induces $2^6 = 64$ possible unique agreement patterns. We note that linkage quality is bound to be quite weak for many potential matches with such a large number of records and such limited identifying information.

@@

Were a personal computer able to hold the 16 GB data object required by the direct approach, the Gibbs sampler under \texttt{BRL} would be tortuously slow, with computatonal complexity $O(n_A \times n_B \times F)$ where $n_A = 20485$. However, under \texttt{fabl} the sampler has complexity $O(P \times n_B \times F)$ where $P = 64$, so 1000 iterations of the Gibbs sampler are completed in just XX seconds. Posterior analysis yields an initial estimate of 9991 individuals linked across the the two datasets, and 8895 individuals linked once conficts are resolved to ensure one-to-one matching.

Lastly, we comment on the significance of conducting this linkage without the use of deterministic blocking. With previous methods, one could link these two records by blocking on a certain field, and running separate Gibbs samplers on each of the seperate blocks. This strategy drastically improves computation since it significantly reduces the number of comparison vectors that need to be formed and uses smaller data objects throughout the sampler. However, this computational savings comes at the cost of missing potential links, specifically in cases where the blocking variable is recorded in error. In this case, 62 of the matched pairs exhibit a disagreement in sex, 17 exhibit a disagreement in state, and 39 exhibit a disagreement in regional office. We have no way to verify that the matches found under \texttt{fabl} are indeed correct, so it is possible that matches found with such disagreements are false positives. Even if not, the modeller may conclude that, for example, 17 missed matches are a small price to pay to be able to block on state and run 50 smaller, independent linkage procedures. However, \texttt{fabl} allows the modeler to make this choice for themselves, rather than being forced to do so for computational reasons. Additionally, by running on the Gibbs sampler the entirety of both files, we achieve $m$ and $u$ parameters that are interpretale for the entire study, rather than separate parameters for each block. Although these $m$ and $u$'s are often regarded as nuisance parameters in the record linkage literature, \texttt{fabl} allows for study of these parameters that has been been possible before.

@@

\begin{center}
{\large\bf ACKNOWLEDGEMENTS}
\end{center}

>>>>

\title{Efficient and Scalable Entity Resolution with Fast Beta Linkage  (fabl)}
\author{Brian Kundinger \and
  Jerome Reiter \and 
  Rebecca C.~Steorts}
\date{
  Department of Statistical Science, Duke University\\
  \today}

@@

In this paper, we propose an extension to the \emph{Beta Record Linkage}
(\texttt{BRL}) method proposed by \citep{sadinle2017} for this very end.  We increase the speed of the \texttt{BRL} procedure through the modification to the model specification that allows for parallel computing of the linkage parameter, and use hashing techniques to hasten calculations and reduce computational complexity. Additionally, we introduce storage efficient indexing (SEI), and a data partitioning approach to make the method amenable to large linkage tasks. To recognize the lineage from the original \texttt{BRL} method, we name our method \emph{Fast Beta Linkage} (\texttt{fabl},
pronounced ``fable'')

@@

\paragraph{Fellegi and Sunter Method}
Most record linkage techniques are derived from the seminal \citep{fs} paper "A Theory for Record Linkage". The defining
characteristic of their approach was to transform the sets of records,
which often contain text data that is difficult to model, into sets of
comparison vectors governed by parameters that can be more easily
estimated. Concretely, if files \(A\) and \(B\) have \(n_A\) and \(n_B\)
records respectively, and if the files share \(F\) fields in common upon
which to base the linkage, the Fellegi and Sunter approach generates an
\(n_A n_B \times F\) matrix \(\Gamma\), which contains similarity scores
between each pair of records across datasets. We say \(\gamma_{ij}\) is
the comparison vector for record \(i \in A\) and record \(j \in B\),
with \(\gamma_{ij}^f \in \{1, \ldots, L_f\}\) providing their similarity
score on the \(f^{th}\) field. For ease of modeling and computation, we
restrict these similarity scores to be discrete, ordinal variables, and
the construction of these is left to the modeler. We adopt the
convention that \(\gamma_{ij}^f = 1\) corresponds to the highest level
of agreement and \(\gamma_{ij}^f = L_f\) corresponds to the lowest. It
is common to use use binary 1-2 variables to indicate exact matching,
and 1-2-3 variables to provide an option for partial matching. For text
data, we calculate similarity based on Levenstein distance or some other
text similarity score, and bin these scores to integers for use in the
model. \citep{jaro1989} provided guidelines for appropriate binning for the
comparison vectors, but details of implementation are generally context
specific.

The likelihood in the original Fellegi Sunter model is a mixture model
through which each record pair is independently classified as a match or
nonmatch. This independent classification often leads to sets of matches
that break expectations of transitivity, which is undesirable. Although
\citep{jaro1995} devised an procedure to reduce a set of conflicting matches to the
mostly likely set of one-to-one matchings, the abundance of false
matches can bias the estimation of other parameters of the model.

@@

\paragraph{Other Approaches (Will probably remove)}Outside of the Fellegi-Sunter framework, some researchers have developed
methods that model the data directly, not a derived set of comparison
vectors. As an early example,  produced
\texttt{blink}, which had the advantage of a computational complexity
that grew linearly with the data (as opposed to quadratically within
Fellegi Sunter), but was plagued with slow mixing times and was unable
to incorporate text data. Later, Steorts et al 2020 improved on this
method with \texttt{dblink}, which was able to simulate text data within
its Gibbs sampler by drawing from an empirical distribution, and
remedied its slow mixing through a probabilistic blocking approach.
However, this method remains computationally intensive and requires
distributed computing to manage even moderate linkage problems in
acceptable time. Because these methods fall outside of the
Fellegi-Sunter framework, we do not consider them further in this paper.

@@

Here the reader should the relationship between our proposed model and
that of Sadinle 2017. In his model, Sadinle constructs a prior
distribution on the entire \(\mathbf{Z}\) vector, which induces a Gibbs
sampler that strictly enforces one-to-one matching. In particular, this
sampler removes previously matches records from the set of candidate
records when sampling \(Z_j\), creating a dependency that makes the
sampler \emph{inherently serial}. We however use independent priors for
each \(Z_j\), creating a sampler that is \emph{perfectly parallelizable}, allowing for significant computational gains. More
importantly, since only the agreement pattern of \(Z_j\) is used for
calculations within the Gibbs sampler, and not the particular record
label, we can conduct this sampling only at the level of the unique
agreement patterns, offering even more computational savings. In doing
so however, we thereby weaken the one-to-one requirement from
\texttt{BRL}; our sampler does ensure that each record in \(B\) can be
matched to at most one record in \(A\), but allows for the possibility
that multiple records in \(B\) match to the same record in \(A\). We
explore the ramifications of this distinction with the El Salvador case
study.

@@

In order to make fair comparisons against the \citep{sadinle2017} model, we
integrate over the posterior of \(\lambda\) and rearrange terms to
produce the final full conditional:

@@

For a more indepth explanation of this function and the induced Bayes
estimate, see \citep{sadinle2017}.

@@

\section{Efficient and Scalable Implementation}
\label{sec:efficiency}
Broadly speaking, we increase our computational efficiency by
recognizing that record pairs contribute to posterior calculations only
through the agreement pattern of the \(\gamma_{ij}\) vector. Let
\(\mathcal{H}\) be the set of unique agreement patterns in the data, let
\(P = |\mathcal{H}|\) denote the total number of unique agreement
patterns. Note that \(P\) is bounded above by \(\prod_{f=1}^F L_f\), and
that this bound does not scale with \(n_A\) or \(n_B\). Prior to
processing the data, we identify all \(P\) patterns in \(\mathcal{H}\)
and enumerate them \(h_1, \ldots, h_P\). When the \((i,j)\) pair
exhibits the \(p^{th}\) agreement pattern, we say \((i,j) \in h_p\).
Wherever possible, we conduct calculations over these \(P\) agreement
patterns rather than the \(n_A \times n_B\) record pairs. Some of these techniques were used by \citep{enamorado2019} to produce \texttt{fastlink} a fast a scalable implementation of the Fellegi-Sunter model, and others have been created here for our particular context.

@@

The classic Fellegi Sunter method represents the \(\gamma_{ij}\)
comparison vector as vector of length \(F\), with each component
\(\gamma_{ij}^f\) taking on values in \(\{1, \ldots, L_f \}\). To ease
computations, we instead represent the comparison as a concatenation of
\(F\) many binary indicator vectors of lengths \(L_f\). For example, if
\(L_1 = L_2 = 2\) and \(L_3 = 3\), then \(\gamma_{ij} = (2, 1, 3)\)
under the classical framework becomes
\(\gamma_{ij} = (0, 1, 1, 0, 0, 0, 1)\) under our framework. This is a
bijective transformation that does not change the meaning of the data,
but this representation will ease calculations and posterior updates. This is also the form the data takes in the \texttt{BRL} package in \texttt{R}.

@@

Lastly, for large data, we can partition the two datasets \(A\) and
\(B\) into smaller blocks \(\{A_m\}\) and \(\{B_m\}\) for more
manageable computations. On a single machine, we can read-in data
sequentially, conduct hashing, compress information through SEI, and
delete the original data from memory before continuing with the next
chunk of data. With multiple cores or multiple machines, this can be
done in parallel. Thus the combination of hashing, SEI, and partitioning
allows us to conduct linkage tasks over much larger datasets.

@@

Lastly, we recognize that all posterior updates are governed by the
agreement patterns of the record pairs rather than the record labels
themselves. Thus we complete the entire Gibbs procedure first at the
level of the \(P\) agreement patterns with the first equation above.
After, we can back-fill the records corresponding to the agreement
patterns through the second equation. Sampling uniformly is
computationally simple even for large sets of candidate records, but this
step can also be parallelized when working with large data.

@@

\begin{figure}
	\[\text{Recall} = \frac{\text{Matches Correctly Identified}}{\text{True Matches in Data}}\]
	\[\text{Precision} = \frac{\text{Declared Matches}}{\text{True Matches in Data}}\]
	\[\text{F-Measure} = 2\left(\frac{\text{Recall} \times \text{Precision}}{\text{Recall} + \text{Precision}}\right)\]

	\caption{Definitions of performance metrics}\label{fig:performance_metrics}
\end{figure}

@@

		\includegraphics[width=0.6\textwidth]{../notes/figures/sadinle_sim_plot} 
	}

@@

To demonstrate speed, we generate comparison vectors from pre-specified
distributions so that we can easily increase the size of the linkage
problem. Distributions are meant to emulate the behavior of similarity
scores across first name, last name, and day, month. For simplicity, we
consider only exact matching, so a vector (1, 0) corresponds to
agreement and (0, 1) to disagreement. We simulate these data for
different values of \(n_A\) and \(n_B\), and compare the run-time of
\texttt{fabl} against \texttt{BRL}. Note that the number of unique
patterns \(P\) is bounded above by \(2^5 = 32\), a bound which is
consistently attained in the larger simulations.

@@

\begin{table}[t]
	\centering
	\begin{tabular}{rrr}
		\hline
		& m & u \\ 
		\hline
		fname & (0.95, 0.05) & (0.01, 0.99) \\ 
		lname & (0.95, 0.05) & (0.01, 0.99) \\ 
		day & (0.95, 0.05) & (0.03, 0.97) \\ 
		month & (0.95, 0.05) & (0.08, 0.92) \\ 
		year & (0.95, 0.05) & (0.07, 0.93) \\ 
		\hline
	\end{tabular}
\caption{Distributions used for $m$ and $u$ probabilties in simulation studies}\label{fig:distributions}
\end{table}

@@

	{\centering \includegraphics[width=0.6\textwidth]{../notes/figures/sadinle_speed_plot2}

@@

	\caption{Run-time for BRL and fabl to run 1000 Gibbs iterations, including hashing step for fabl, for increasing values of both $n_A$ and $n_B$}\label{fig:speed1}
\end{figure}

@@

	{\centering \includegraphics[width=0.6\textwidth]{../notes/figures/speed_plot_fixed_nB}

@@

	\caption{Run-time for BRL and fabl to run 1000 Gibbs iterations, including hashing step for fabl, with increasing $n_A$, and $n_B$ fixed at 500}\label{fig:speed2}
\end{figure}

@@

We now demonstrate the power of our method through two case studies. The first is a study of documented identifiable deaths (DID) from the El Salvadoran Civil War. Though the data files used here are small, this study shows how the computational complexity of \texttt{fabl} depends on the number of unique agreement patterns, and how significant computational gains can be achieved by simplifying the construction of the comparison vectors. Secondly, the case study reveals the impact of independently sampling $Z_j$ rather than strictly enforcing one-to-one matching as done by \texttt{BRL}. The second case study is the National Long Term Care Survey (NLTCS), a much larger linkage tasks which \texttt{BRL} is unable to complete but which we complete with \texttt{fabl} with relative ease.

\subsection{El Salvadoran Civil War}
\label{el_salvador}

The country of El Salvador was immersed in civil war from 1980 to 1991,
and throughout the time, several organizations attempted to document
casualties of the conflict. When estimating the total number of
casualties, one cannot simply sum the numbers recorded by each
organization, as it is likely that the same individuals are recorded in
multiple casualty lists. To obtain a more accurate estimate of the
casualties then, we follow Sadinle's 2017 paper and link records of El
Salvadoran civilian casualties from two sources: El Rescate - Tutela
Regal (ERTL) and the Salvadoran Human Rights Commission (CDHES, by its
acronym in Spanish). The ERTL dataset consists of digitized reports that
had been published throughout the conflict. The CDHES dataset consists
of casualties that had been reported directly to the organization, and
later digitized.

@@

We only utilized records with nonmissing entries for given and last
name, results in \(n_A = 4420\) files in CHDES and \(n_B = 1323\) files
in ERTL. The names were standardized to account for common misspellings
in the Spanish language, and then compared using a modified Levenstein
distance to account for the fact that second names are often omitted.
Place of birth is recorded by municipality and department within that
municipality; however, since department was missing in 95\% of records
in CHDES and 80\% of records in ERTL, we excluded it from our linkage
process. Thus we conduct linkage using given name, last name,
municipality, and day, month, and year of death. We again use flat
priors for the \(\mathbf{m}\) and \(\mathbf{u}\) parameters.

@@

	{\centering \includegraphics[width=0.6\textwidth]{../notes/figures/el_salvador/m_posterior_smallP}

@@

To mirror the original implementation, we constructed the comparison
vectors using 4 levels of agreement for each field, according to the
thresholds provided in Figure XX. This took 422 seconds for our proposed
method and 240 for \texttt{BRL}. However, we observed that posterior
distributions of several levels of the \(\mathbf{m}\) and \(\mathbf{u}\)
parameters were nearly identical, and that of the
\(4^5 \times 2 = 2048\) possible agreement patterns, only 1173 are
realized in the data. This leads us to believe that such high number of
agreement levels creates unnecessary distinctions in the data and makes
the comparison vectors less interpretable. Therefore we re-ran our
analysis with fewer agreement levels for each field (see Figure XX), and
obtained analogous results. With 216 possible agreement patterns, 159
were realized in the data, and our proposed method became much faster,
finishing in 124 seconds. Meanwhile \texttt{BRL} took 239 seconds,
relatively unchanged from the first implementation. This demonstrates
the way that the computational complexity of our method depends on the
number of unique agreement patterns, and how significant computational
gains can be made by simplifying the construction of the comparison
vectors. We also note that estimates of the \(\mathbf{m}\) and
\(\mathbf{u}\) parameters under each method are very similar, as shown in
Figure~\ref{fig:m-and-u}.

\paragraph{Violations of One-to-One Matching} Figure~\ref{fig:overlap-plot} shows the posterior distribution for \(D\), the number of
duplicates found across file for both \texttt{fabl} and \texttt{BRL}. We
see that \texttt{fabl} consistently overmatches within each Gibbs
iteration when compared to \texttt{BRL}, which is to be expected because
\texttt{BRL} explicitly prevents matches that violate one-to-one
matching throughout the entire sampler. Most of such matchings are due
to the randomness in the sampling procedure, and they occur sporadically
throughout the sampler in such a way that does not measurably influence
the eventual Bayes estimate.

@@

	{\centering \includegraphics[width=0.6\textwidth]{../notes/figures/el_salvador/overlap_distribution_smallP_bayes}

@@

Overall, both methods presented similar results. \texttt{fabl} yielded
an initial Bayes estimate of 195 matches found across files, and after
resolving matches that violated one-to-one requirements, yielded 180
matches across files. This is acceptably close to the 181 matches found
by \texttt{BRL}. The main reason for this discrepancy is the difference
in how each model handles situations in which one file in \(B\) has
multiple plausible matches in \(A\).

@@

	{\centering \includegraphics[width=0.6\textwidth]{../notes/figures/el_salvador/rosa_records}

@@

	{\centering \includegraphics[width=0.6\textwidth]{../notes/figures/el_salvador/bad_mixing}

@@

Figure~\ref{fig:mixing-plot} shows the values that \(Z_{825}\) and \(Z_{826}\) take on
throughout the Gibbs sampler, and demonstrates how each method handles
this situation. Under \texttt{fabl}, both records in \(B\) match to the
same record in \(A\) throughout the Gibbs process, creating consistent
violations of one-to-one matching. Under \texttt{BRL}, the Gibbs process
creates one matching configuration stays there for a while. However, if
one pair ``unmatches,'' then the other record has a chance to latch on.
Then, the Gibbs process is stuck with that matching status for a while,
resulting in a Gibbs process with poor mixing. Additionally,
\texttt{fabl} allows the modeler to inspect records with multiple
plausible matches, and if they desire, to then choose the record pairing
with the highest posterior probability. \texttt{BRL} in contrast, in
strictly enforcing one-to-one matching throughout the sampler, can lead
to situations where none of the plausible matches reach the threshold to
be identified through the Bayes estimate.

\subsection{National Long Term Care Survey}
\label{nltcs}

The National Long Term Care Survey is a longitudinal study to track changes in health among Medicare recipients. It was conducted approximately every five years, and we concern ourselves here with linking records from years 1982 and 1989. The data only contains day, month, and year of birth, sex, state the individual lives in, and regional office. State and regional office have been anonymized, presented as numbers instead of their actual values. We construct the comparison vectors using only binary exact matching for each field, since there is no straightfoward way to incorporate partial matching in this context. The 1982 file contains $n_A = 20485$ records, 1989 file contains $n_B = 17466$ files, and the six linkage fields induces $2^6 = 64$ possible unique agreement patterns. We note that linkage quality is bound to be quite weak for many potential matches with such a large number of records and such limited identifying information.

@@

Were a personal computer able to hold the 16 GB data object required by the direct approach, the Gibbs sampler under \texttt{BRL} would be tortuously slow, with computational complexity $O(n_A \times n_B \times F)$ where $n_A = 20485$. However, under \texttt{fabl} the sampler has complexity $O(P \times n_B \times F)$ where $P = 64$, so 1000 iterations of the Gibbs sampler are completed in just XX seconds. Posterior analysis yields an initial estimate of 9991 individuals linked across the the two datasets, and 8895 individuals linked once conflicts are resolved to ensure one-to-one matching.

Lastly, we comment on the significance of conducting this linkage without the use of deterministic blocking. With previous methods, one could link these two records by blocking on a certain field, and running separate Gibbs samplers on each of the separate blocks. This strategy drastically improves computation since it significantly reduces the number of comparison vectors that need to be formed and uses smaller data objects throughout the sampler. However, this computational savings comes at the cost of missing potential links, specifically in cases where the blocking variable is recorded in error. In this case, 62 of the matched pairs exhibit a disagreement in sex, 17 exhibit a disagreement in state, and 39 exhibit a disagreement in regional office. We have no way to verify that the matches found under \texttt{fabl} are indeed correct, so it is possible that matches found with such disagreements are false positives. Even if not, the modeler may conclude that, for example, 17 missed matches are a small price to pay to be able to block on state and run 50 smaller, independent linkage procedures. However, \texttt{fabl} allows the modeler to make this choice for themselves, rather than being forced to do so for computational reasons. Additionally, by running on the Gibbs sampler the entirety of both files, we achieve $m$ and $u$ parameters that are interpretale for the entire study, rather than separate parameters for each block. Although these $m$ and $u$'s are often regarded as nuisance parameters in the record linkage literature, \texttt{fabl} allows for study of these parameters that has been been possible before.

@@

\begin{center}
{\large\bf ACKNOWLEDGMENTS}

Left over from dblink paper, might remove.
\end{center}
